\input{thinkos-common.tex}

% Define Title Author Date and Revision
\makeatletter         
\title{ThinkOS - User Guide}
\author{Bob Mittmann}
\date{January 2017}
\def\email{bobmittmann@gmail.com} % revision number variable
\def\therevision{0.2} % revision number variable
\let\thetitle\@title % title variable
\makeatother

\begin{document}

\maketitle
\thispagestyle{empty}
\newpage
\pagenumbering{Roman}
\tableofcontents
\newpage
\listoffigures
\newpage
\listoftables
\newpage

% Start of the revision history table
\begin{versionhistory}
  \vhEntry{0.1}{21.01.17}{BM}{created}
  \vhEntry{0.2}{22.03.18}{BM}{patially fixed redundant sections}
\end{versionhistory}

\newpage
\pagenumbering{arabic}

\section{Introduction}

\lettrine{T}{his} user guide provides an overview of the \ThinkOS real-time operating system. The target audience for this guide are embedded software developers knowledgeable in C language. Notice that no mention of C++ is made in this text.\\

\ThinkOS was designed specifically for the \textbf{Cortex-M} family of microprocessors from \textbf{ARM}. These are all 32bits CPUs with mostly 16bits instruction set, which produces very compact code, hence suitable for deep embedded systems with stringent memory and other peripheral resources constraints.

At this moment \ThinkOS supports \textbf{M3}, \textbf{M4} and \textbf{M7} cores.

The popular \textbf{M0} family lacks some features needed for some operations in the kernel. Porting to these platforms is feasible but needs someone with time to do that. Any volunteer?

ThinkOS is a Real Time Operating System designed for the ARM Cortex-M core. It takes advantage of some unique features of this processor to enable for very low task switch latency.


\subsection{Why Another OS?}

The real answer for this question perhaps is simply: ``Just because I can''. But as most people won't be satisfied with this shallow reply here are some justifications:

\subsection{Why Cortex-M}

The ARM Cortex-M family of microprocessors is the standard 32 bits CPU for most of the microcontrollers in the market today.

\subsection{Can it be ported to another processor?}

In theory yes. But probably its not worth the effort...


\subsection{Where do I can get ThinkOS from?}

\begin{flushleft}
Repository: [\href{https://github.com/bobmittmann/thinkos}{https://github.com/bobmittmann/thinkos}]\\
Git HTTPS: [\href{https://github.com/bobmittmann/thinkos.git}{https://github.com/bobmittmann/thinkos.git}]
\end{flushleft}

\subsection{How to use it?}
 How and what is need in order to compile \ThinkOS ....
\begin{itemize}
\item Download the source code
\item Install the Tools
\item Configure / or chose and example
\item Compile
\item Load and run
\end{itemize}


\subsection{Performance}

The main bottleneck for performance is usually the scheduler. The ThinkOS running in a 120MHz Cortex-M3 (STM32F270) have a measured latency time of 0.5 microseconds.

There are some tests and calculations that need to be done to determine the latency of other elements of the system...


\subsection {Source Code}

One feature that distinguishes the \ThinkOS from most of other Operating Systems is the way the control structures for the several types of objects are held by kernel.

The usual approach for holding this information is to allocate a structure specific for a certain object and pass a pointer to the kernel. The kernel then stores this in a list or array. 

\ThinkOS in other hand holds the core data structure in a  Structure of Arrays (SOA) representation. The cost of this representation is the readability of the code, (it may look a little obscure at the beginning). 
The idea here is simplify the race condition avoidance in the kernel and speed-up the scheduler.

\subsubsection{Coding Style}

\ThinkOS source uses the \href{https://www.kernel.org/doc/html/v4.10/process/coding-style.html}{Linux kernel coding style} and I encourage all developers to follow it. This is just little more than a matter of preference, please read the style guide to know the rational behind the rules before despising them :) ...

%----------------------------------------------------------
%== Introduction ==
%ThinkOS is a Real Time Operating System designed for the ARM Cortex-M core. It takes advantage of some unique features of this processor to enable for very low task switch latency.

\subsection{Compiling}

There are basically two ways of using \ThinkOS:

\paragraph{As a library}{ linked against your application. (Need a term to decribe this like embedded mode...). 
In this type of usage there is usually no bootloader, the application boots and configure itself. In this scenario the firmware is the application. The advatnages of this mode are:}

\begin{itemize}
\item Smallest footprint. The linker will strip away unused (unreferenced) portions of the kernel.
\item Kernel calls and services are tailored for the application.
\end{itemize}
	
\paragraph{As part of the bootloader}{ separated from the application. This mode will be called "standalone" mode, or "bootloader" mode.
In this case the application will be a separated entity (file, binary) and run in a separated memory space. The bootloader will be residing on the platform at all times and will be responsible to initialize the platform and to load the application. It's possible then to upload upgrade and change the application without changing the bootlader and consequently the operating system. Advantages:}

\begin{itemize}
\item No need to configure the OS (need to elaborate...) to match the application. At least initially.
\item In field firmware upgrades are safer as the bootloader is the responsible for the upgrade.
\item The Debug Monitor features can be used for development.
\item Kernel data and code memory are protected.
\end{itemize}

% \subsubsection{System Calls}
\subsection{Modules}
 Overview of the different modules.
\begin{itemize}
\item Kernel
\item Debug Monitor
\item GDB server
\item Console
\end{itemize}


\subsection{Threads}

\ThinkOS is a preemptive time-sharing kernel. Each thread has its own stack and a timer. This timer is used to block the thread's execution for a certain period of time as in \lstinline{thinkos_sleep()}.

\subsection{Scheduler}
ThinkOS scheduler was designed to reduce the context switch time to a minimum. The most important piece on the Kernel is the Scheduler....

\subsection{Time Sharing}
All threads with priority higher than 0 go into a round-robin time sharing scheduling policy, when this feature is enabled. Their execution time will be inversely proportional to their priority.

\subsection{Interrupt Model}
There are two types of interrupt handlers with ThinkOS:

\paragraph{Implicit ISR}{Thread interrupt handlers that executes as normal threads. This is a simple call to \lstinline{thinkos_irq_wait(IRQ)} inside a regular task or function. The thread will block until this interrupt is raised. This mechanism provides a convenient and powerful tool to design drivers. Devices that benefit most of this type of ISR are simplex I/O channels like I2C or SPI, hardware accelerators like cryptographic ...}

\paragraph{Native ISR}{Normal interrupt handlers that shares a common stack. These handlers are subject to to some constraints regarding API calls. It is recommended to use this type of interrupt handler when low latency is a must. But bear in mind that if the handler is too complex the actual latency due to loading registers and housekeeping may be equivalent to a context switch in which case it will be equivalent of a threaded handler in terms of performance. Another case will be a high priority interrupt you want to make sure it will won't be preempted...}


Configuration is simply a matter of creating your custom configuration header file: \lstinline{config.h} and enabling or disabling the features the application requires.

The file \lstinline{include/thinkos_sys.h} contains the definition of the structures used by the OS. In this file there is a series of macros used to set default values for the configuration options.

It is not recommended to change the values directly in \lstinline{thinkos_sys.h}, create a \lstinline{config.h} file instead, and define the macro \lstinline{CONFIG_H} at the compiler's: \lstinline{-DCONFIG_H}.

Note: if you are using the \ThinkOS as part of the YARD-ICE there is a \lstinline{config.h} file located at the source root.


% ----------------------------------------------------------------------------

\section {Quick Start}

This guide will help you to install the necessary tools to compile and run a ThinkOS demo application on a development board.

\subsection{Windows Tools}

These are the basic steps to follow in order to install the tools: 

\begin{itemize}
\item Install GNU Toolchain for ARM Processros for Windows hosts.
\item Install GNU Make for Windows.
\item Install GNU CoreUtils for Windows.
\item Install the Java JRE from Oracle.
\item Install the Eclipse for C++ Developers IDE.
\item Install the GNU ARM Eclipse Plug-ins.
\item Install the C/C++ GDB Hardware Debugging Plug-in.
\item Install the STM32 ST-LINK Utility from STMicroelectronics.
\end{itemize}

Detailed instructions for each one of the steps can be found in the next sections of this document. 

\subsection{Demo Application}
There are some demo applications using \ThinkOS in the source tree. This is a guide on how to compile and run ThinkOS programs in the STM32F3Discovery board. 


\begin{itemize}
\item Get the \ThinkOS Source Code from GitHub.
\item Import the Eclipse projects for the target board into your workspace.
\item Compile the \ThinkOS BootLoader project using Eclipse IDE.
\item Load the \ThinkOS BootLoader on the board using the STM32 ST-LINK Utility.
\item Install the Virtual COM Driver on the PC to access the ThinkOS Debug-Monitor.
\item Compile the SDK Libraries (sdklib) for the target board using Eclipse.
\item Compile the demo application project using Eclipse.
\item Run the application with GDB Remote Debugging.
\end{itemize}



\subsection{ThinkOS Basics}

\lstset {language=C99}

In this tutorial the most basic \ThinkOS functions are described with some code examples. This is a summary of what it will be covered in this section:

\begin{itemize}
\item How to create and run threads.
\item Create, lock and unlock mutexes.
\item Create, signal and wait on semaphores.
\item Wait for time intervals (sleep).
\item Create a new thread
\end{itemize}

\subsection{Creating a Thread}

The easiest way to create and run a thread is to call \lstinline{thinkos_thread_create()}. Here is an example: 

\begin{lstlisting}[caption={Simple Thread Create Example},label={lst:example1},language=C99,frame=single]
static int my_task(void * arg)
{
    for (;;) {
        /* do something ... */
         thinkos_sleep(50);
    }
    return 0;
}

static uint32_t my_stack[256]; /* 1KB of stack is enough for most porposes */

void my_module_init(void)
{
    thinkos_thread_create(my_task, NULL, my_stack, sizeof(my_stack));
}
\end{lstlisting}

In this example we define a function to be the entry point of the thread, in this case \lstinline{my_task()}. We also define an array of 32bit words, amounting to 1KB, to be used as stack for the thread (\lstinline{my_stack}). We then call \lstinline{thinkos_thread_create()} which will allocate and run the thread. The return of this function is the thread id or a negative error code in case something goes wrong.

\subsubsection{Using semaphores}

ThinkOS semaphores allow processes and threads to synchronize their actions.

A semaphore is an integer whose value is never allowed to fall below zero. Two operations can be performed on semaphores: increment the semaphore value by one (\lstinline{thinkos_sem_post()}); and decrement the semaphore value by one (\lstinline{thinkos_sem_wait()}). If the value of a semaphore is currently zero, then a \lstinline{thinkos_sem_wait()} operation will block until the value becomes greater than zero. A semaphore should be allocated with \lstinline{thinkos_sem_alloc()} which will return a descriptor to be used in subsequent calls to  \lstinline{thinkos_sem_post()}, and \lstinline{thinkos_sem_wait()}.

In the example below, two treads are created: a producer and a consumer. The producer signals the semaphore periodically twice a second, whereas the consumer waits for the signal and prints a message when receives it.

\begin{lstlisting}[caption={Using semaphores example},label={lst:example2},language=C99,frame=single]
/* Global semaphore descriptor */

int my_sem;

/* Producer thread */
static int producer_task(void * arg)
{
    for (;;) {
        /* Wait 500 milliseconds... */
        thinkos_sleep(500);
        /* Signal the semaphore. */
        thinkos_sem_post(my_sem);
    }
    return 0;
}

/* Consumer thread */
static int consumer_task(void * arg)
{
    for (;;) {
        /* Wait on the semaphore... */
        thinkos_sem_wait(my_sem);
        /* Do something... */
        printf("Signal received\n");
    }
    return 0;
}


static uint32_t consumer_stack[256];
static uint32_t producer_stack[256];

void my_test_init(void)
{
    /* Allocate a semaphore with 0 initial count. */
    my_sem = thinkos_sem_alloc(0);
    /* Create a thread to signal the semaphore */
    thinkos_thread_create(producer_task, NULL, producer_stack, sizeof(producer_stack));
    /* Create a thread to wait on the semaphore */
    thinkos_thread_create(consumer_task, NULL, consumer_stack, sizeof(consumer_stack));
}
\end{lstlisting}

\subsubsection{Using MUTEXes}

A mutex is a MUTual EXclusion device, and is useful for protecting shared data structures from concurrent modifications, and implementing critical sections and monitors.

A mutex has two possible states: unlocked (not owned by any thread), and locked (owned by one thread). A mutex can never be owned by two different threads simultaneously. A thread attempting to lock a mutex that is already locked by another thread is suspended until the owning thread unlocks the mutex first.

A call to \lstinline{thinkos_mutex_lock()} locks the given mutex. If the mutex is currently unlocked, it becomes locked and owned by the calling thread, and \lstinline{thinkos_mutex_lock()} returns immediately. If the mutex is already locked by another thread, \lstinline{thinkos_mutex_lock()} suspends the calling thread until the mutex is unlocked by calling \lstinline{thinkos_mutex_unlock()}. A mutex should be allocated with \lstinline{thinkos_mutex_alloc()} which will return a descriptor to be used in subsequent mutex calls.

In the following example the serial port is a resource that needs to be protected to ensure the packets are transmitted in full.


\begin{lstlisting}[caption={Using MutExes examples},label={lst:example3},language=C99,frame=single]
/* Link Layer private data */
static struct {
	uint8_t addr; /* this panel address */
	
	
    struct serial_dev * serial; /* serial device driver */
    int mutex; /* protect the link layer send call */
} rs485;

void netlnk_send(unsigned int daddr, void * data, unsigned int len)
{
    struct lnkhdr hdr;

    /* lock the mutex, blocking the access to the serial port... */
    thinkos_mutex_lock(rs485.mutex);
    /* prepare header */
    hdr.sync = PKT_SYNC;
    hdr.daddr = daddr;
    hdr.saddr = rs485.addr;
    hdr.datalen = len;
    /* send header */
    serial_send(rs485.serial, &hdr, sizeof(struct lnkhdr));
    /* send payload */
    serial_send(rs485.serial, data, len);
    /* unlock mutex, other thread can use the serial port now. */
    thinkos_mutex_unlock(rs485.mutex);
}

void netlnk_init(unsigned int addr)
{
    /* Open the serial port */
    rs485.serial = stm32f_uart1_serial_init(9600, SERIAL_8N1);
    /* Set the local address */
    rs485.addr = addr;
    /* Allocate a mutex */
    rs485.mutex = thinkos_mutex_alloc();
}
\end{lstlisting}

\subsubsection{Time Wait}

Use the \lstinline{thinkos_sleep()} function to wait for a period of time. The calling thread suspends its execution until at least the time specified as parameter has elapsed. The time period is specified in milliseconds.
The example that follows a LED will blink repeatedly 4 times per second:

\begin{lstlisting}[caption={Time Wait Example},label={lst:example4},language=C99,frame=single]
int blink_task(void * arg)
{
    for (;;) {
        led_on();           /* Turn LED on */
        thinkos_sleep(100); /* Wait for 100 milliseconds */
        led_off();          /* Turn LED off */
        thinkos_sleep(150); /* Wait for 150 milliseconds */
    }

    return 0;
}
\end{lstlisting}

\subsubsection{Compiling the Code}

In order to compile the code in a Windows machine you must first install these tools:
\begin{itemize}
\item GNU Toolchain for ARM Processors for Windows hosts.
\item GNU Make for Windows.
\end{itemize}

\subsubsection{Compiling Using the Command Shell}

You can use the windows Command Prompt or a MinGW/MSYS terminal to compile the code.
Open a windows shell: Start->All Programs->Accessories->Command Prompt
Change to the directory  in the ThinkOS source tree.
Type: make

\section{\ThinkOS Kernel}

 - Data structures.
 
 - Design decisions.
 
 ...

\subsection{Exception Priorities}



\subsection{Kernel Objects}

\subsubsection{System Calls}

\subsubsection{Library functions}


%\subsection{Configuring}
% Description on how to configure ThinkOS.
% - bootloader mode
% - embedded mode


\section{API}

The ThinkOS API can be separated according to the type of object the operation is performed ...

\paragraph{Core}{initialization configuration and debug}

\paragraph{Threads}{functions related to threads}

\paragraph{Mutexes}

\paragraph{Conditional Variables}

\paragraph{Semaphores}

\paragraph{Events}

\paragraph{Interrupts}

\paragraph{Time}


\input{thinkos-api.tex}

\section {Demo Using Windows}

%\input{thinkos-demo-win.tex}


\section {Tools}
% = SECTION ===================================================================
\section{Tools}

% == SUB-SECTION =========================================
\subsection{MSYS-2}

MSYS2 is software distribution and a building platform for Windows. It provides a Unix-like environment, a command-line interface and a software repository making it easier to install, use, build and port software on Windows.\\

The official website is:

\href{http://www.msys2.org/}{MSYS2 Website}\\

Useful information can be found at the project's wiki pages at: 

\href{https://github.com/msys2/msys2/wiki/MSYS2-introduction}{MSYS2 Introduction}\\

% == SUB-SECTION =========================================
\subsection{GNU Toolchain for ARM Processors}

ThinnkOS on Windows Host - this document is a guide on how to set-up the tools and environment to develop ThinkOS applications using a Windows computer as host............

The GNU toolchain is a broad collection of programming tools developed by the GNU Project. These tools form a toolchain (a suite of tools used in a serial manner) used for developing software applications and operating systems.

If you are a windows user, you can download and install the GNU Toolchain for ARM Processors from the site: https://launchpad.net/gcc-arm-embedded. At the time of writing this document the latest version known to work with YARD-ICE was 4.9-2015-q3, which can be downloaded directly from:

\begin{flushleft}
GNU Arm Embedded Toolchain at Launchpad:
[\href{https://launchpad.net/gcc-arm-embedded}{https://launchpad.net/gcc-arm-embedded}]\\
\end{flushleft}

\begin{flushleft}
[\href{https://developer.arm.com/open-source/gnu-toolchain/gnu-rm}{GNU Arm Embedded Toolchain}]\\

Downloads: [\href{https://developer.arm.com/open-source/gnu-toolchain/gnu-rm/downloads}{https://developer.arm.com/open-source/gnu-toolchain/gnu-rm/downloads}\\
\end{flushleft}

Run the downloaded file \textit{gcc-arm-none-eabi-6-2017-q2-update-win32-sha2.exe}, or newer, to start the installation.
In the next steps use the default values suggested by the wizard installer.
In the last screen, after the files were copied, select the "Add path to environment variable" option.

Alternatively a compressed (.zip) file can be downloaded (\textit{gcc-arm-none-eabi-6-2017-q2-update-win32.zip}) and extracted in a directory of choice...

\subsubsection{GNU Make}

\subsubsection{GNU Core Utils for Windows}

% == SUB-SECTION =========================================
\subsection{Python}

% == SUB-SECTION =========================================
\subsection{Julia}


% == SUB-SECTION =========================================
\subsection{Eclipse}

% == SUB-SECTION =========================================
\subsection{Tera-Term}


% == SUB-SECTION =========================================
\subsubsection{DCCLog}

\section{Apendix}

\section{Efficient Developer}

\subsection{Master your tools}

\subsection{Automated Code Generation}

Embedded systems ...
In some circumstances there are some pieces of code can be ...
better performance comprises of a repetition of code that is tedious to type and error prone. Also in some contexts a generic solution that can be computed at runtime can be too expensive in terms of CPU usage to be computed in real-time. Or may require too much memory. A common solution for these problems is to use lookup tables and to interpolate intermidiate values, store coefficients in FLASH.
Even if you could afford to calculate a complex dataset some systems require a very fast boot-up response, so storing pre computed values is a sure way of impproving things.
We present here two typical scenarios where an automated generated code is benneficial for embedded systems.

\subsubsection{Fixed-point sine function}

Example sine function using Python.


\subsubsection{IIR filter}

In this example we use a Julia script to calculate and automatically generate a very efficient C code for a low pass filter. By automating the code generation its easy to change the filter parameters, like cur off frequency, filter order without having to worry about synchronizing the design specs with your embedded code.

\subsubsection{Embedded HTML pages}

Here HTML pages, or any pages for that matter, are automatically compressed, converted into C arrays. Makefile rules take care of the dependency of your code and the web pages, so the code is regenerated and compiled whenever a page is modified.


\subsubsection{Autogenerated Code Comments}

Always add a comment in the code indicating that it is automatically generated and a summary of instructions on how to regenerate it.

Example:

\subsection{Makefiles}

\subsection{Text Editor}

\subsubsection{Multiple editor windows}

\subsection{Git - Revision Control}

\section {References}

\subsection {Gnu Tools for Window}

More information on the GNU Make and CoreUtils for Windows package can be found at:

\begin{itemize}
\tightlist
\item
  {\href{http://gnuwin32.sourceforge.net/packages/make.html}{http://gnuwin32.sourceforge.net/packages/make.htm}}
\item
  {\href{http://gnuwin32.sourceforge.net/packages/coreutils.htm}{http://gnuwin32.sourceforge.net/packages/coreutils.htm}}
\end{itemize}

\subsection{GNU ARM Eclipse Plug-Ins}

{Quick info on GNU ARM Eclipse Plug-Ins:}

\begin{itemize}
\tightlist
\item
  {Name: }{`\emph{GNU ARM Eclipse Plug-ins}'}
\item
  {Location:
  }{\href{http://gnuarmeclipse.sourceforge.net/updates}{http://gnuarmeclipse.sourceforge.net/updates}}
\item
  {Website:
  }{\href{http://gnuarmeclipse.livius.net/blog}{http://gnuarmeclipse.livius.net/blog}}
\end{itemize}


\subsection{GNU MCU Eclipse plug-ins}

These plug-ins provide Eclipse CDT (C/C++ Development Tooling) extensions for GNU embedded toolchains like the GNU Tools for ARM Embedded Processors used as reference in this document.

\begin{itemize}
\tightlist
\item
	{Eclipse update site:
	}{\href{http://gnu-mcu-eclipse.netlify.com/v4-neon-updates}{http://gnu-mcu-eclipse.netlify.com/v4-neon-updates}}
\item
	{Website:
	}{\href{https://gnu-mcu-eclipse.github.io}{https://gnu-mcu-eclipse.github.io}}
\end{itemize}

\subsection{Git Extensions}

From the project's website: \quote{Git Extensions is a toolkit aimed at making working with Git under Windows more intuitive (note that Git Extensions is also available on Linux and Macintosh OS X using Mono). The shell extension will integrate in Windows Explorer and presents a context menu on files and directories.}\\


\paragraph{Website: }\href{http://git-extensions-documentation.readthedocs.io/en/latest/git_extensions.html}{Git Extensions}


% == SUB-SECTION =========================================
\subsection{YARD-ICE}

The YARD-ICE project is a community based, open source hardware and software platform for a standalone remote debugger. The embedded platform uses \ThinkOS as bootloader and real-time operating system.

\begin{itemize}
\tightlist
\item
  {YARD-ICE Project Location:
  }{\href{https://github.com/bobmittmann/yard-ice}{https://github.com/bobmittmann/yard-ice}}
\item
  {Latest Release:
  }{\href{https://github.com/bobmittmann/yard-ice/archive/0.24.zip}{https://github.com/bobmittmann/yard-ice/archive/0.24.zip}}
\item
  {Git Repository (HTTPS):}
  {\href{https://github.com/bobmittmann/yard-ice.git}{https://github.com/bobmittmann/yard-ice.git}}
\item
  {Git Repository (SSH):}{\href{git@github.com:bobmittmann/yard-ice.git}{git@github.com:bobmittmann/yard-ice.git}}
\end{itemize}

\subsection{Useful Web References}

\begin{flushleft}
\href{https://en.wikipedia.org/wiki/Real-time_operating_system}{Real-time operating system} - Wikipedia - March, 2017
\end{flushleft}

\begin{flushleft}
\href{https://www.gnu.org/gnu/thegnuproject.en.html}{The GNU Project}\\
\end{flushleft}

\begin{flushleft}
\href{https://gcc.gnu.org/}{GCC, the GNU Compiler Collection}\\
\end{flushleft}

\begin{flushleft}
\href{https://github.com/msys2/msys2/wiki}{MSYS2}\\
\end{flushleft}

\begin{flushleft}
\href{https://git-scm.com/downloads}{Git}\\
\end{flushleft}

\begin{flushleft}
\href{https://www.kernel.org/doc/html/v4.10/process/coding-style.html}{Linux kernel coding style}\\
\end{flushleft}


\end{document}

